---
title: "ABM for testing spatial memory"
output: html_document
---

##Introduction:

This is a notebook that will impliment an ABM to test assumptions of spatial memory on travel decisions. The code was writen for the paper "Chimpanzees use advanced spatial cognition to plan least-cost routes to out of sight goals" in Current Biology.

Outline:

0. get the data ready
1. Define the ABM
2. Simulate potential paths using the ABM
3. Plot the results

### 0) Get the data

Load libraries
```{r}
library(gdistance)
library(ggplot2)
library(raster)
library(rgdal)
library(circular)
library(movecost)
library(dplyr)
library(doParallel)
```

Add spatial data
```{r}
#DEM
str_name <- "SRTM_UTM_Clip1.tif"
r <- raster(str_name)
crs(r)

#Paths
str_name_paths <- "EstabTrail_Cost21.tif"
paths <- raster(str_name_paths)
crs(paths)

#start end points
df.start.end.coords <- read.csv("StartEnd_Coords.csv")
df.start.end.coords.starts <- df.start.end.coords %>% dplyr::select(Start.x, Start.y)
df.start.end.coords.ends <- df.start.end.coords %>% dplyr::select(End.x, End.y)

coordinates(df.start.end.coords.starts) <- ~ Start.x + Start.y
coordinates(df.start.end.coords.ends) <- ~ End.x + End.y

crs(df.start.end.coords.starts) <- "+proj=utm +zone=35 +south +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0"
crs(df.start.end.coords.ends)   <- "+proj=utm +zone=35 +south +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0"

#look at it all
plot(r)
plot(paths, add =T, alpha=0.35)
points(df.start.end.coords.starts, col="red")
points(df.start.end.coords.ends, col="blue")

```


### 2) Define the ABM

```{r}
###################
#
# Functions used to calculate costs
#
#################

#function to convert cost values to conductance values
cost.to.connectance <- function(x){
  x.min=min(x)
  x.max=max(x)
  x2<-(1 - ((x - x.min) / (x.max - x.min))) * (x.max - x.min) + x.min
  return(x2)
}

#function to convert cost values to conductance values: 1/cost
cost.to.connectance.inv <- function(x){
  return(1/x)
}

vert.factor.slope <- function(x){
  
  #convert to degrees
  slope.degree = (x * 180) / (pi)
  
  vf <- ifelse(slope.degree < -15, 0.64, ifelse(slope.degree < 0,1+slope.degree*0.024 ,1+slope.degree*0.05 ))
  
return(vf)  
}

#mean difference between two cells
cost.fun.path <- function(x){
  return((x[2]+x[1])/2)
}

#difference between two cells
heightDiff <- function(x){x[2] - x[1]}

#cost distance function
cost.distance <- function(x){
  
  #get difference in elevation
  elevation.diff = x[2] - x[1]
  tot.dist <- sqrt( (elevation.diff^2) + (30.88^2) )
  return(tot.dist)
  
}



###################
#
# Functions used to calculate transitions
#
#################


#calculates the conductance surface
getConductance_reduced <- function(start, end){
  
  #cut the rasters to smaller extent
  min_x = min(start[1],end[1])
  max_x = max(start[1],end[1])
  min_y = min(start[2],end[2])
  max_y = max(start[2],end[2])
  
  #set extent
  r.temp <- raster()
  bb <- extent(min_x - 2000,  max_x + 2000,min_y - 2000, max_y + 2000)
  extent(r.temp) <- bb
  res(r.temp) <-res(r)
  projection(r.temp) <- crs("+proj=utm +zone=35 +south +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
  
  #clip
  r.clip<-crop(x=r, y=r.temp)
  paths.clip<-crop(x=paths, y=r.temp)#extent.temp
  
  ###################
  #
  # Elevation
  #
  #################
  
  hd <- transition(r.clip,heightDiff,8,symm=FALSE)
  slope <- geoCorrection(hd, scl=FALSE)
  adj <- adjacent(r.clip, cells=1:ncell(r.clip), pairs=TRUE, directions=8)
  vf.cost <- slope
  vf.cost[adj] <- vert.factor.slope(x=slope[adj])
  vf.cost[adj] <- cost.to.connectance.inv(x=vf.cost[adj])
  
  #############
  #
  # Paths
  #
  #############
  
  #calculate mean conductance between cells based on path/non-path of cells
  paths.cost <- transition(paths.clip, cost.fun.path, 8, symm=FALSE)
  adj <- adjacent(paths.cost, cells=1:ncell(paths.cost), pairs=TRUE, directions=8)
  paths.cost[adj] <- cost.to.connectance.inv(x=paths.cost[adj])
  
    #################
  #   
  #  Combine
  #
  #################
  
  #combined cost of travel 
  comb <-   paths.cost * vf.cost 

  #correct for distance between cells
  comb<-geoCorrection(comb, scl=F)
  
  return(comb)
  
}

#calculates the conductance surface
getConductance_full <- function(){
  
  ###################
  #
  # Elevation
  #
  #################
  
  hd <- transition(r,heightDiff,8,symm=FALSE)
  slope <- geoCorrection(hd, scl=FALSE)
  adj <- adjacent(r, cells=1:ncell(r), pairs=TRUE, directions=8)
  vf.cost <- slope
  vf.cost[adj] <- vert.factor.slope(x=slope[adj])
  vf.cost[adj] <- cost.to.connectance.inv(x=vf.cost[adj])
  
  #############
  #
  # Paths
  #
  #############
  
  #calculate mean conductance between cells based on path/non-path of cells
  paths.cost <- transition(paths, cost.fun.path, 8, symm=FALSE)
  adj <- adjacent(paths.cost, cells=1:ncell(paths.cost), pairs=TRUE, directions=8)
  paths.cost[adj] <- cost.to.connectance.inv(x=paths.cost[adj])
  
    #################
  #   
  #  Combine
  #
  #################
  
  #combined cost of travel 
  comb <-   paths.cost * vf.cost 

  #correct for distance between cells
  comb<-geoCorrection(comb, scl=F)
  
  
  return(comb)
  
  
  
}


###################
#
# Functions used to simulate movement
#
#################


#main function controling the simulation
simulate.abm<-function(x){
  out.sim.final <- data.frame(x=-1, y=-1, total_cost=-1, total_dist=-1 ,madeit=-1,rowID=-1,rep=-1)
  for(rowV in x){
    
    #get start and end points
    start = c(df.start.end.coords[rowV,]$Start.x,df.start.end.coords[rowV,]$Start.y)
    end   = c(df.start.end.coords[rowV,]$End.x,df.start.end.coords[rowV,]$End.y)
    
    #get conductance surface
    costSurf_sub <- getConductance_reduced(start, end)
    
    for(rep in  1: sim.replicates){
      
      #sim
      df.temp <- simulate.random.path(start, end,costSurf_sub,rowV,rep)
      
      #record
      out.sim.final<-rbind(out.sim.final,df.temp)
    }
    
    print(paste0("stepped ",rowV))
    
  }
  if(nrow(out.sim.final)>1)out.sim.final<-out.sim.final[-1,]
  return(out.sim.final)
}


#function for calculating agent path
simulate.random.path <- function(start, end, costSurf,x,rep){
  
  #Agent variables
  total_dist = 0
  total_cost = 0
  current_position = c(start[1],start[2])
  madeit = 0
  
  #inialize a dataframe
  out.sim.final <- data.frame(x=current_position[1], y=current_position[2], total_cost=total_cost, total_dist=total_dist ,madeit=madeit,rowID=x,rep=rep)
  
  #simulate steps
  for(i in 1:sim.steps){
    
    #make a move
    #calculate neighbouring points
    step.results <- calculate.next.move(current_position, costSurf = costSurf)
    
    #update agent
    total_cost <- total_cost + step.results[[2]]
    total_dist <- total_dist + (((step.results[[1]][1]-current_position[1])^2) + ((step.results[[1]][2]-current_position[2])^2))^0.5
    current_position <- step.results[[1]]
    
    #check to see if the agent reached the end
    if(calculate.distance.betw.points(current_position,end) < min.threshold ){
      
      madeit=1
      
      #make the last step
      step.results <- calculate.final.move(current_position, end , costSurf = costSurf)
    
      #update agent
      total_cost <- total_cost + step.results[[2]]
      total_dist <- total_dist + (((step.results[[1]][1]-current_position[1])^2) + ((step.results[[1]][2]-current_position[2])^2))^0.5
      current_position <- step.results[[1]]
      
      #record each step
      out.sim.final <- data.frame(x=current_position[1], y=current_position[2], total_cost=total_cost, total_dist=total_dist ,madeit=madeit,rowID=x,rep=rep)
      
      #end the simulation
      break
      
    } 
    
    #record each step
    out.sim.final <- rbind(out.sim.final,data.frame(x=current_position[1], y=current_position[2], total_cost=total_cost, total_dist=total_dist ,madeit=madeit,rowID=x,rep=rep))
    
  }
  
  #after all steps, record the results
  if(madeit==0){
    #out.sim.final <- data.frame(x=start[1], y=start[2], total_cost=total_cost, total_dist=total_dist ,madeit=0)
  }
  
  return(out.sim.final)
  
}


#####calculate potential local paths within a visual radius, and choose the best one.
calculate.next.move <- function (currentCell, costSurf){
  
  #calculate neighbouring points
  random.points <- raster::adjacent(costSurf, cellFromXY(costSurf,xy=c((currentCell)[1],  (currentCell)[2]) ) ,directions=8)
  random.points<-t(apply( as.matrix(random.points[,2]) , 1,get_xy,r=costSurf ))
  
  #calculate the path cost of each local option
  random.points.cost <- apply(random.points,1,calculate.cost.of.route, start=currentCell, costSurf=costSurf)
  chosenIndex <- sample(1:nrow(random.points), prob=1/random.points.cost, size =1)

  #return new updated start location and the cost to tavel to it
  return(list(random.points[chosenIndex,],random.points.cost[chosenIndex]) )
  
}

#####calculate the final path to the end point.
calculate.final.move <- function (currentCell,end, costSurf){
  
  #calculate the path cost of each local option
  end.points.cost <- costDistance(x=costSurf, fromCoords=currentCell, toCoords=t(as.matrix(end)) )
  
  #return new updated start location and the cost to tavel to it
  return(list(end,end.points.cost) )
  
}


get_xy <- function(x, r){
  xyFromCell(r,x)  
}

calculate.points.x.meters.away <- function(start, x, vision){
  return(c(start[1]+cos(x)*vision,start[2]+sin(x)*vision))
  
}

calculate.cost.of.route <- function(x, start,costSurf){
  return(costDistance(x=costSurf, fromCoords=start, toCoords=x))
} 


calculate.distance.betw.points <- function(x,p2){
  return((((x[1]-p2[1])^2) + ((x[2]-p2[2])^2) )^0.5)
}


```


#### 3) run the ABM

Choose parameters for the ABM (all parameters for the ABM will be placed here).
```{r}
#df.start.end.coords <- read.csv("StartEnd_Coords.csv")
#df.start.end.coords<-df.start.end.coords[,1:6]

min.threshold = 45     #when has the agent reached the end point
sim.replicates = 100   #How many simulated replicates to run
sim.steps = 122        #How many steps does the agent take (maximum)

```


Run the abm in paralle
```{r}

#Create the clusters
cl <- makeCluster(10)  
registerDoParallel(cl)

#Export the environment variables to each cluster
clusterExport(cl,ls())

#Load the library "rgeos" to each cluster
clusterEvalQ(cl, library(rgeos))
clusterEvalQ(cl, library(raster))
clusterEvalQ(cl, library(gdistance))

#Split the data
ID.Split<-clusterSplit(cl,1:nrow(df.start.end.coords)) #

#Run the function in each cluster
system.time(m<-clusterApply(cl, x=ID.Split, fun=simulate.abm)) #ID.Split

#Cluster close
stopCluster(cl)

#Merge the results
output<- do.call("rbind", m)

#output the results to a csv file
write.csv(output, file="SimpathCosts.csv")
```


#### 4) Look at the outputs

Plot results
```{r}
results1 <-output

#create unique id column
results1$unID <- paste0(results1$rowID,"_", results1$rep)

#summarise the chance of making it for each rowID 
results1.summarised <- results1 %>% group_by(unID) %>% summarise(madeit.final = max(madeit), rowID=max(rowID), distanceTraveled = max(total_dist), costTotal = max(total_cost))
results1.chance <- results1.summarised %>% group_by(rowID) %>% summarise(chance = sum(madeit.final)/n(), )  

#Add rowID to the observed start/end coordinates
df.start.end.coords.join <- df.start.end.coords 
df.start.end.coords.join$rowID <- seq(1:nrow(df.start.end.coords.join)) 

#bind the observed and the simulated dataframes
results2<-left_join(results1.chance,df.start.end.coords.join, by="rowID")

#create unique id
results2$ID = paste0(round(results2$Start.x),"_",round(results2$Start.y)," ",round(results2$End.x),"_",round(results2$End.y))

#take a look
ggplot(results2, aes(x=reorder(factor(ID), chance), y=chance)) + geom_point() + theme_classic() + theme(axis.text.x = element_blank()) + geom_hline(yintercept=0.5, linetype = "dashed") + labs(y="Probability of reaching the end point", x="Travel segment")
```



